Project Review Factors Influencing Abundance & Virus Isolation in Mosquitoes

### 2024-04-29 R/analysis-code/analysis.qmd
Debug and clean
debugged code and updated README files

# Load the necessary libraries for the analysis

```{r}
library(here)
library(dplyr)
library(lubridate)
library(readr)
library(stargazer)
library(mgcv)
library(tidyr)
library(caret)
library(randomForest)
library(tidymodels)
library(ranger)
library(dials)
library(vip)
library(ggplot2)
library(boot)
library(yardstick)
library(glmnet) # For LASSO
library(dials)
library(mlr3)
library(data.table)
library(Matrix)
library(gbm)
library(knitr)
library(gt)
```

@@

```{r}
pcr <- read_csv(here("data", "processed-data", "pcr_data.csv"))
wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))

@@

# tuning_grid <- grid_regular(
#   mtry(range = c(1, length(predictors))),
#   min_n(range = c(1, 21)),
#   levels = 7
# )

@@

# Calculate the bootstrapped RMSE
mean_rmse_boot_rf <- mean(bootstrap_results$t)
sd_rmse_boot_rf <- sd(bootstrap_results$t)
ci_rmse_boot_rf <- boot.ci(bootstrap_results, type = "perc")  # Percentile interval

@@

autoplot(tuning_results_continuous)
```
### The random forest model has evidence of overfitting as the RMSE on the test set is higher than the training set.

@@

autoplot(tuning_results_continuous)

>>>>

# Load the necessary libraries for the analysis
```{r}
library(here)
library(dplyr)         # for data manipulation tasks like filtering, mutating, summarizing, and arranging
library(lubridate)     # for working with dates and times, offering functions to parse, manipulate, and format date-time objects
library(readr)         # for reading and writing rectangular data in text format, providing faster and more consistent reading of delimited files
library(stargazer)     # for creating LaTeX and ASCII tables for descriptive statistics and regression models
library(mgcv)          # for generalized additive models (GAMs) and other generalized ridge regression models
library(tidyr)         # for tidying and reshaping data for analysis and visualization
library(caret)         # for machine learning tasks like model training, tuning, and evaluation
library(randomForest)  # for building random forest models, a popular machine learning algorithm
library(tidymodels)    # for modeling and machine learning tasks using tidy principles
library(ranger)        # for fast, robust, and scalable random forest algorithms
library(dials)         # for tuning hyperparameters in models
library(vip)           # for variable importance assessment in predictive models
library(ggplot2)       # for creating elegant and customizable visualizations with the grammar of graphics
library(boot)          # for bootstrap resampling and estimating standard errors and confidence intervals
library(yardstick)     # for evaluating predictive models using performance metrics
library(glmnet)        # for fitting generalized linear models with LASSO or ridge regularization
library(mlr3)          # for building, evaluating, and comparing machine learning models
library(data.table)    # for efficient data manipulation and aggregation operations
library(Matrix)        # for sparse and dense matrix operations
library(gbm)           # for gradient boosting machines, a powerful machine learning algorithm
library(knitr)         # for dynamic report generation, integrating R code and output into documents
library(gt)            # for creating beautiful and customizable tables in R Markdown and other formats
library(xgboost)       # for extreme gradient boosting, a scalable and accurate machine learning algorithm)

@@

#### This script starts by loading essential R libraries necessary for data manipulation, reading, and date handling, ensuring a smooth setup for data analysis. It then loads PCR test results and weather data, providing foundational information on mosquito test results and environmental conditions to study patterns of disease spread and mosquito activity. Finally, it calculates average maximum temperatures for key months before the mosquito breeding season, linking climatic conditions to mosquito behavior and potential disease transmission risks.
```{r}
# Step 1: load the data
pcr <- read_csv(here("data", "processed-data", "pcr_data.csv"))
wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))

@@

@@

boot_fit_and_evaluate <- function(data, indices) {
  # Subset the bootstrap sample
  boot_sample <- data[indices, ]

  # Predict on the bootstrap sample using the final fitted model
  predictions <- predict(final_fit_continuous, new_data = boot_sample)

  # Ensure predictions are in the correct format, typically need a vector
  # If predictions are not a vector, adjust accordingly
  if (is.data.frame(predictions)) {
    predictions <- predictions$.pred  # Adjust this based on your predict function output structure
  }

  # Calculate the RMSE of predictions
  # Ensure both truth and predictions are either integers or doubles
  results <- tibble(
    truth = as.numeric(boot_sample$MIR),  # Convert to numeric to ensure matching types
    estimate = as.numeric(predictions)  # Convert to numeric to match type
  )

  # Compute RMSE and return
  result <- rmse(results, truth = truth, estimate = estimate)
  return(result$.estimate)
}
# Continue with bootstrapping to evaluate model's robustness
set.seed(123) # for reproducibility
bootstrap_results <- boot(
  data = train_data_continuous,
  statistic = boot_fit_and_evaluate,
  R = 100 # number of bootstrap resamples
)
# Calculate the bootstrapped RMSE
mean_rmse_boot_rf <- mean(bootstrap_results$t)
sd_rmse_boot_rf <- sd(bootstrap_results$t)
ci_rmse_boot_rf <- boot.ci(bootstrap_results, type = "perc")  # Percentile interval

@@

plot <- autoplot(tuning_results_continuous)

# Save the plot to the specified directory
ggsave(filename = here("results", "figures", "RF_autoplot.png"), plot = plot, width = 10, height = 8)
```
### The random forest model has evidence of overfitting as the RMSE on the test set is higher than the training set.

@@

plotGBM <- autoplot(tuning_results_continuous)

#save plot as a png
ggsave(filename = here("results", "figures", "GBM_autoplot.png"), plot = plotGBM, width = 10, height = 8)

### 2024-04-29 R/eda-code/EDA_unified.qmd
Debug and clean
debugged code and updated README files

# load needed packages. make sure they are installed.
```{r}
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(readr)
library(forecast)
library(tsibble) # For time series data wrangling
library(feasts) # For decomposing and visualizing time series
library(RColorBrewer)
library(ggcorrplot)
library(readxl)
library(corrplot)
library(plotly)
library(reshape2)
library(rlang)
```

@@

# Join the MIR data with ither datasets based on week number
#### First we will process some data that will be used in the join
```{r}
# Load the datasets:
MIR_by_week_co2 <- read_csv(here("data", "processed-data", "MIR_by_week_co2.csv"))
wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
mosquito_wide <- read_csv(here("data", "processed-data", "mosquito_wide_all.csv"))
drought_fema1_4 <- read_csv(here("data", "processed-data", "drought_fema1_4.csv"))
drought_ma_counties <- read_csv(here("data", "processed-data", "drought_ma_counties.csv"))
average_daylight_by_week <- read_csv(here("data", "processed-data", "average_daylight_by_week.csv"))
average_mel_weekco2 <- read_csv(here("data", "processed-data", "average_mel_weekco2.csv"))
wx_hourly_clean <- read_csv(here("data", "processed-data", "hourly_wx_clean.csv"))
mel_prop_weekly <- read_csv(here("data", "processed-data", "mel_proportion_by_week.csv"))

>>>>

# load needed packages. make sure they are installed.
```{r}
library(here)        # for data loading/saving
library(dplyr)       # for data manipulation
library(skimr)       # for data summarization
library(ggplot2)     # for data visualization
library(lubridate)   # for date and time manipulation
library(forecast)    # for time series forecasting
library(tsibble)     # for time series data wrangling
library(feasts)      # for decomposing and visualizing time series
library(RColorBrewer) # for color palettes
library(ggcorrplot)  # for correlation plot
library(readxl)      # for reading Excel files
library(corrplot)    # for correlation plot
library(plotly)      # for interactive plots
library(tidyr)       # for data reshaping
library(reshape2)
```

@@

# Join the MIR data with ither datasets based on week number
#### First we will process some data that will be used in the join
```{r}
# Load the datasets:
MIR_by_week_co2 <- read_csv(here("data", "processed-data", "MIR_by_week_co2.csv"))
wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
mosquito_wide <- read_csv(here("data", "processed-data", "mosquito_wide_all.csv"))
drought_fema1_4 <- read_csv(here("data", "processed-data", "drought_fema1_4.csv"))
drought_ma_counties <- read_csv(here("data", "processed-data", "drought_ma_counties.csv"))
average_daylight_by_week <- read_csv(here("data", "processed-data", "average_daylight_by_week.csv"))
average_mel_weekco2 <- read_csv(here("data", "processed-data", "average_mel_week_co2.csv"))
wx_hourly_clean <- read_csv(here("data", "processed-data", "hourly_wx_clean.csv"))
mel_prop_weekly <- read_csv(here("data", "processed-data", "mel_proportion_by_week.csv"))

### 2024-04-29 R/processing-code/MADAproject_pt2_ruiz.qmd
Debug and clean
debugged code and updated README files

# Loading data
## Mosquito trap data
### This dataset contains trap counts by species for all mosquito traps set in the Bristol County from 2007-2023 this dataset will undergo extensive processing below.
#### The dataset was obtained though and open records request with the Commonwealth of MA.
```{r trap-data-import}
# Construct the path to the Excel file
file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")
file_exists_check <- file.exists(file_path_mos)
file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")

@@

# Save the trap_location_selected dataframe as an RDS file
saveRDS(trap_location_selected, trap_location_selected_path_rds)
```

@@

### The data includes all the positive PCR results for EEE and WNV for all mosquitoes submitted for testing
### These files were obtained from: https://www.mass.gov/lists/arbovirus-surveillance-plan-and-historical-data
### The virus isolation data is stored in separate spreadsheets by year from 2014 to 2020
### This code will load all the files, combine them, and then save the combined file in the "raw data" folder
```{r}
#Define the path to the folder with Excel files
folder_path <- here("data", "processed-data", "virus_data")

@@

## Weather data
### temperature and precipitation are important factors in mosquito population dynamics
### This dataset contains daily weather summaries from the Taunton, MA Airport from 2007-2023
#### The data was obtained from the NOAA National Centers for Environmental Information
```{r}
# Construct the path to the file
file_path_wx = here("data", "processed-data", "weather_airport.csv")

# import the data to R and store it in a dataframe named "wx"
wx = read.csv(file_path_wx)

wx <- wx %>%
  mutate(
    DATE = mdy(DATE),  # Parse the DATE column from "month/day/year"
    week_num = paste0(year(DATE), "-", sprintf("%02d", isoweek(DATE)))  # Create week_num
  )
#load the first few rows of data to verify the data loaded as expected
head(wx)

@@

#### The majority of the missing data is in the "Submitted for Testing" variable. This is not a crucial variable, so we can keep those records. Missing data in that variable generally means 0 mosquitoes were captured in a trap. The other variable are important and so we will remove those records with the other missing data.

### The Massachussetts Department of Public Health adpoted a taxonomy change before it was fully vetted. The change divided the genus *Aedes* into a new genus: *Ochlerotatus*. To align with accepted standards, I will change them back to *Aedes*.

@@

### Execute proposed changes.

@@

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)

@@

```{r annual precip}
# Convert DATE from character to Date type using as.Date and specifying the format
wx$DATE <- as.Date(wx$DATE, format="%m/%d/%y")

# The as.Date function treats years in the format "yy" as follows:
# Years 00-68 are treated as 2000-2068, and years 69-99 are treated as 1969-1999.
# Adjust this according to your data's actual year range if needed.

# Extract the year from each DATE
wx$Year <- format(wx$DATE, "%Y") %>% as.numeric()

@@

>>>>

# Loading data
## Mosquito trap data
### This dataset contains trap counts by species for all mosquito traps set in the Bristol County from 2007-2023 this dataset will undergo extensive processing below.
#### The dataset was obtained though and open records request with the Commonwealth of MA.
##### we will read the data now and process it later in this script
```{r trap-data-import}
# Construct the path to the Excel file
file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")
file_exists_check <- file.exists(file_path_mos)
file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")

@@

# Save the trap_location_selected dataframe as an RDS file
saveRDS(trap_location_selected, trap_location_selected_path_rds)

#save as csv
trap_location_selected_path_csv <- here("data", "processed-data", "trap_sites_selected.csv")
write_csv(trap_location_selected, trap_location_selected_path_csv)

@@

### The data includes all the positive PCR results for EEE and WNV for all mosquitoes submitted for testing
### These files were obtained from: https://www.mass.gov/lists/arbovirus-surveillance-plan-and-historical-data
### The virus isolation data is stored in separate spreadsheets by year from 2014 to 2020
### This code will load all the files, combine them, and then save the combined file in the "raw data" folder
```{r}
#Define the path to the folder with Excel files
folder_path <- here("data", "raw-data", "virus_data")

@@

## Weather data
### temperature and precipitation are important factors in mosquito population dynamics
### This dataset contains daily weather summaries from the Taunton, MA Airport from 2007-2023
#### The data was obtained from the NOAA National Centers for Environmental Information
```{r}

# Construct the path to the file
file_path_wx <- here("data", "raw-data", "weather_airport.csv")

# Import the data to R and store it in a dataframe named "wx"
wx <- read.csv(file_path_wx)

# Check the structure of the DATE column to determine if there are any unexpected values
str(wx$DATE)

# Check for any missing values in the DATE column
sum(is.na(wx$DATE))

# If there are no missing values, proceed with the mutation
wx <- wx %>%
  mutate(
    DATE = ymd(DATE),  # Parse the DATE column from "year-month-day"
    week_num = paste0(year(DATE), "-", sprintf("%02d", isoweek(DATE)))  # Create week_num
  )

# Load the first few rows of data to verify the data loaded as expected
head(wx)

@@

#### The majority of the missing data is in the "Submitted for Testing" variable. This is not a crucial variable, so we can keep those records. Missing data in that variable generally means 0 mosquitoes were captured in a trap. We will remove the rows with missing data in the other columns in th code chunks below.

### The Massachussetts Department of Public Health adpoted a taxonomy change before it was fully vetted. The change divided the genus *Aedes* into a new genus: *Ochlerotatus*. To align with current accepted standards, I will change them back to *Aedes*.

@@

### Execute proposed from above

@@

@@

```{r annual precip}
# Construct the path to the file
file_path_wx <- here("data", "processed-data", "wx_selected.csv")

# Import the data to R and store it in a dataframe named "wx"
wx <- read.csv(file_path_wx)

# Extract the year from each DATE and store it in a new column named "Year"
wx$Year <- as.integer(format(as.Date(wx$DATE), "%Y"))

@@

# View the filtered dataframe
head(filtered_wx)

### 2024-04-29 peer_reviews/CMiller-project-review-Ruiz.qmd
Debug and clean
debugged code and updated README files

***Personal Message***  
Andrew, this looks great.  From our conversations over Discord, you were very critical of your own work.  This project looks awesome.  You have a couple of things to polish off, but, wow.  You obviosuly have the topic covered, you have the coding skills, you got this.

@@

*Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?*

At this point, this section of the document is incomplete.

@@

At this point, this section of the document is incomplete.  That being said, the description of the intended analysis is appropriate.

@@

There are no results at this time.

@@

At this point, this section of the document is incomplete.

@@

* results are poorly presented, hard to understand, poor quality

At this point, this section of the document is incomplete.

@@

At this point, this section of the document is incomplete.

@@

At this point, this section of the document is incomplete.

### Summary assessment
* major parts of discussion missing or wrong

At this point, this section of the document is incomplete.

@@

A large portion of this work is missing.  It is my understanding that the author is working through these sections again as new data (rainfall/weather data) was made avaiable.

The basic bones of the project look good.

@@

Not at this time.

@@

At this point, this section of the document is incomplete.

@@

Yes.  As far as it goes.  At this point, this section of the document is incomplete.

@@

At this point, this section of the document is incomplete.

@@

Once complete, the described work will be very complete.  At this point this section is incomplete.

@@

At this point, this section of the document is incomplete.

@@

Just finish this project off the way you've described and it will be great.  Hang in there!

>>>>

***Personal Message***  
Andrew, this looks great.  From our conversations over Discord, you were very critical of your own work.  This project looks awesome.  You have a couple of things to polish off, but, wow.  You obviously have the topic covered, you have the coding skills, you got this.

@@

*Were the analysis methods appropriate for the data? Was the analysis done properly?*

Yes, I believe that they were.  I am not an expert on this topic, but the analysis looked reasonable and properly done.

*Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?*

Again, I am not an expert on this topic.  The analysis looked appropriate for the data type.  Yes, the data and analysis were explained well.

@@

No real feedback on this part.  You did a great job in this category.  The description of the intended analysis is appropriate.

@@

The results were presented in a table format.  The results compared several types of model for the desired data.  The table included RMSE data for training and test data.  No figures of the analysis were made.

@@

@@

* results are presented ok, with room for improvement

At this point, this section of the document seems incomplete.  I can see a need for tables describing the output discussed in the basic data analysis section.  Plus, maybe some descriptive figures of the analysis if possible?

@@

Yes.  The conclusions are discussed, the strengths and limitations of the data and analyses are discussed.

@@

No real feedback on this section.  The section looked generally complete.

* strong, complete and clear discussion

@@

The basic bones of the project look good.  It is obvious that a good deal of work has gone into the paper.

@@

Not at this time.  There are still a lot of extra files in the repo.

@@

I think that no project can be documented well enough.  This one needs a bit more work in that area.  There are still a a lot of readme files talking about how the project will look when it's finished.  Of course, that might just be github acting up again.

@@

Yes.  Everything looks complete and it runs.

@@

I like how you piled all the code into a single R folder.  It took a minute for me to find it because it wasn't the format that my other review used.  I think that is a cleaner way to organize the coding sections.

@@

The overall study looks very thorough and complete.  Again, I am not an expert in this area, but everything looked reasonable and well thought out.  All the salient areas seemed to be addressed.

@@

No real feedback on this section.  The section looked generally complete.

@@

Just finish this project off the way you've described and it will be great!
